---
title: "Business-oriented Typology and Characterization of Agriculture in SSA"
author: '[Melanie Bacou](http://github.com/mbacou) for BMGF'
date: "Last updated on `r Sys.Date()`. DRAFT, DO NOT USE OR CITE."
output:
  html_notebook:
    fig_caption: yes
    fig_height: 5
    toc: yes
    toc_depth: 3
    toc_float: yes
link-citations: yes
nocite: |
  @aasr_2016, @alvarez2014typology, @jayne2015agdev, @hazell2013urban, @douillet2014developing, @jordan2005eth, @omamo2006strategic, @bacou2015ethseg, @omamo2006strategic, @benin2016agricultural, @willy2015adaptation, @jayne2015africafarmland
csl: apa.csl
subtitle: A multi-scale and cross-time framework
bibliography: biblio.bib
---

--------------------------------------------------------------------------------------

```{r, message=FALSE}

library(data.table)
library(foreign)
library(stringr)
library(rhandsontable)
library(jsonlite)

load("../tmp/2017-agra-aasr.RData")

```


```{r, eval=FALSE, collapse=TRUE}

# Helper - Construct summary stats
summaryTable <- function(dt) {
  
  dt.lbl <- data.table(
    varCode=names(dt),
    varLabel=attr(dt, "var.labels")
  )  
  tmp <- as.data.table(summary(dt))
  tmp[, V1 := 1:7]
  tmp <- dcast(tmp, V2~V1)
  setnames(tmp, c("varCode", "Min.", "1st Qu.", "Median", "Mean", "3rd Qu.", "Max.", "NA's"))
  tmp[, varCode := str_trim(varCode)]
  
  setkey(tmp, varCode)
  setkey(dt.lbl, varCode)
  dt.lbl <- tmp[dt.lbl]
  dt.lbl <- dt.lbl[names(dt)]
  setcolorder(dt.lbl, c(1,9,2:8))
  
  dt <- data.table(dt)
  dt.lbl[, `Hist.` := sapply(dt, function(x) ifelse(class(x)=="numeric", toJSON(list(values=hist(x, plot=FALSE)$counts)), toJSON(NA)))]
  return(dt.lbl)
}

# Variables extracted for the Poverty, Land, Climate paper
hh1 <- read.dta("../tmp/Final_SSA_hhPOV_ 6 Apr 2016.12.dta")
hh.dim <- list(hh1=dim(hh1))

# Show summaries
hh1.lbl <- summaryTable(hh1)
hh1 <- data.table(hh1)

# List surveys/years included in this set
dt1sum <- rhandsontable(hh1[, .(
  Regions=uniqueN(svyL1Cd), 
  Districts=uniqueN(paste(svyL1Cd, svyL2Cd)), 
  Obs=.N), keyby=.(ISO3, Year=as.integer(year))],
  rowHeaders=NULL)
rm(hh1)

# Describe variables
dt1var <- rhandsontable(hh1.lbl, rowHeaders=NULL) %>%
  hot_col("Hist.", renderer = htmlwidgets::JS("renderSparkline"))

# Load dataset #2
hh2 <- read.dta("../tmp/Combined_Regdemovars13.12.dta")
hh.dim$hh2 <- dim(hh2)

# Show summaries
hh2.lbl <- summaryTable(hh2)
hh2 <- data.table(hh2)

# List surveys/years included in this set
dt2sum <- rhandsontable(hh2[, .(
  Regions=uniqueN(svyL1Cd), 
  Obs=.N), keyby=.(ISO3, Year=as.integer(year))],
  rowHeaders=NULL)
rm(hh2)

# Describe variables
dt2var <- rhandsontable(hh2.lbl, rowHeaders=NULL) %>%
  hot_col("Hist.", renderer = htmlwidgets::JS("renderSparkline"))

# Load dataset #3
hh3 <- read.dta("../tmp/Combined_4_Mel.12.dta")
hh.dim$hh3 <- dim(hh3)

# Show summaries
hh3.lbl <- summaryTable(hh3)
hh3 <- data.table(hh3)

# List surveys/years included in this set
dt3sum <- rhandsontable(hh3[, .(
  Regions=uniqueN(region), 
  Districts=uniqueN(paste(region, district)), 
  Obs=.N), keyby=.(ISO3, Year=as.integer(year), round)],
  rowHeaders=NULL)
rm(hh3)

# Describe variables
dt3var <- rhandsontable(hh3.lbl[1:150], rowHeaders=NULL) %>%
  hot_col("Hist.", renderer = htmlwidgets::JS("renderSparkline"))

# Save workspace
save.image("../tmp/2017-agra-aasr.RData")

```

# Background

This study is a prospective chapter in *AGRA Africa Agriculture Status Report (AASR) 2017*.

The overall objective of the AASR report is to [@aasr_cn_2017]:  
 
1. provide an overview of the smallholder farmers and how they have adapted to the challenges they face as economic actors; 
2. explore innovative strategies that can substantially raise the productivity and incomes of smallholder farmers; 
3. identify policies and programs that can support the movement of Africa’s farming systems from subsistence-oriented to market-oriented thriving businesses; 
4. identify the necessary conditions, appropriate technologies, and institutions that can propel and support smallholder agriculture businesses; 
5. examine the past and the present role of public and private sector investment in agriculture and the success factors that can be scaled up to accelerate transformation.  

# Method

The study should provide a context-setting chapter that attempts to characterize, scale, locate, point to prioritization of high-level smallholder commercialization strategies, opportunities and challenges. A proposed approach is to chracterize SHFs and frame business strategies within a 2-by-2 domain framework (low/high rainfed ag potential across low/high market access areas). This approach assumes that agricultural development and adaptation strategies are largely driven by pre-existing bio-physical and spatial conditions.

We propose then focusing on around half a dozen countries for which we have recent/accessible microdata (e.g. LSMS-ISA and maybe AGRA baseline surveys) to look more closely, in the same 2x2 domain framework, at specific farm household and macro characteristics to do two things:  

1. Apply some typology to distinguish between say “predominantly-subsistence” focused and “transitioning-commercial” smallholders  
2. Report on hh level variables that can provide insight into the scale of potential business development challenges and opportunities in each country. 

Key farm and household-level variables to look at may include:

* Holding size
* Income/consumption
* Access to inputs/extension
* Use of inputs
* Land tenure status
* Education
* Cell-phone ownership
* Cooperative/self-help group membership
* Access to credit, insurance
* Off-farm income
* ...

# Supporting Datasets

Many (unpublished and mostly country-level) studies have attempted to derive farmer characteristics and farm typologies across geospatial dimensions (see e.g. refs below). IFPRI in particular generated cross-country comparable household variables (derived from World Bank LSMS-ISA panels, Demographic and Health Surveys, and/or other large-scale household surveys and agricultural census).

## Cross-Country Harmonized Variables

Inventory and summarize the latest versions of IFPRI's harmonized variables. Some of these variables are also available (summarized) across districts and/or regions and/or gender.

### Dataset #1 (Poverty, Land, Climate paper)

This dataset contains `r hh.dim$hh1[2]` variables and `r hh.dim$hh1[1]` observations. The breakdown across countries and survey years is as follows, and variables are further described in the next table. Many of these variables are also document in [@azzarri2016poverty].

```{r}

dt1sum

```
      

```{r}

dt1var

```

### Dataset #2 (Farm variables for Ag Snapshots)

The household-level file is corrupt (would need to contact IFPRI directly) but the variables are similar in the region-level file.

This dataset contains `r hh.dim$hh2[2]` variables. The breakdown across countries and survey years is as follows, and variables are further described in the next table.

```{r}

dt2sum

```


```{r}

dt2var

```


### Dataset #3 (Panels for Resilience study)


This dataset contains `r hh.dim$hh3[2]` variables and `r hh.dim$hh3[2]`  observations. The breakdown across countries and survey years is as follows, and variables are further described in the next table.

```{r}

dt3sum

```

This dataset includes over 2,000 constructed climatic variables (SPEI, temperature, rainfall, PDSI, etc.). They are not shown in the variable summary below.

```{r}

dt3var

```

## Demographic and Health Surveys (harmonized)


## Other Country-Level Panel Variables



# References

Note that some of these references are unpublished materials, not to be cited.


